### 2017-10-16 Draft2/Appending.Dynamic.tex
Rename some stuff

I will first implement adding for dynamic arrays. Let $\listname$ be a dynamic array. The following definitions are used in the code:

\begin{description}
	\item[initial capacity] The capacity of a dynamic a array when one item is added to it. I denote this $\initcapacity$.\\
	(The capacity of any dynamic array with size zero should be zero.)
	\item[growth factor] The factor by which the current capacity is multiplied to get the new capacity when $\listname$ is resized. I denote this with $\growthfactor$.
\end{description}

@@

Before I analyze the time complexity of [add], I consider a different method for measuring its cost. Suppose I start with an empty list and $\varnitems$ elements are added. How many times is an element stored in an array? I will term the answer to this question the \textbf{write cost} of $\varnitems$ adds, and denote it $\nwritesfn(n)$.

In the above code, one write is performed after the conditional, so it is apparent that $\nwritesfn(\varnitems) \geq \varnitems$ after $n$ adds. However, [resize] also does some writing, so in order to find a precise formula for $\nwritesfn(\varnitems)$ I need to analyze when [resize] is called. Consider the following lemma:

[lemma]
Let $\listname$ by a dynamic array. As $\varnitems$ are added, the sequence of capacities $\listname$ takes on is $$S_C = 0,\ \initcapacity,\ \growthfactor\initcapacity,\ \growthfactor^2\initcapacity,\ \ldots\ \growthfactor^{\lceil \log_{\growthfactor} \varnitems - \log_{\growthfactor} \initcapacity \rceil}\initcapacity$$.

[proof]
The capacity starts out at zero and becomes $\initcapacity$ once one element is added. From then on, it can only grow by a factor of $\growthfactor$ via [resize], so if $\growthfactor^i\initcapacity$ is the current capacity then $\growthfactor^{i + 1}\initcapacity$ must be the next capacity. By induction, the rest of the sequence is $$\left\{ \growthfactor^i\initcapacity \right\}_{i = 1}^k$$ for some $k$. Since the capacity is only as big as is needs to be, namely greater than or equal to $\varnitems$, $k$ is the smallest integer for which $\growthfactor^k\initcapacity \geq \varnitems$. Using this property, I now derive $k$.

@@

Each time [resize] is called, $i$ items are copied where $i$ is the current size of the dynamic array. Then the total number of items copied for $\varnitems$ adds is

@@

Finally, adding the writes made for every item in [add], I finally find an explicit formula for $\nwritesfn(\varnitems)$.

@@

I wish to find the space allocated while adding $\varnitems$ items to a dynamic array. I denote this quantity $\spacefn(\varnitems)$, and define it as the total length of all arrays allocated during $\varnitems$ [add] calls. I now derive a formula for $\spacefn(\varnitems)$.

From [lemma] I know the exact capacities $\listname$ takes on as $\varnitems$ items are added. Dynamic arrays also have the property that $i \in S_C$ iff an array of length $i$ was allocated. Then the total length of arrays allocated is

>>>>

I will first implement appending for dynamic arrays. Let $\listname$ be a dynamic array. The following definitions are used in the code:

\begin{description}
	\item[initial capacity] The capacity of a dynamic a array when one item is appended to it. I denote this $\initcapacity$.\\
	(The capacity of any dynamic array with size zero should be zero.)
	\item[growth factor] The factor by which the current capacity is multiplied to get the new capacity when $\listname$ is resized. I denote this with $\growthfactor$.
\end{description}

@@

Before I analyze the time complexity of [append], I consider a different method for measuring its cost. Suppose I start with an empty list and $\varnitems$ elements are appended. How many times is an element stored in an array? I will term the answer to this question the \textbf{write cost} of $\varnitems$ appends, and denote it $\nwritesfn(n)$.

In the above code, one write is performed after the conditional, so it is apparent that $\nwritesfn(\varnitems) \geq \varnitems$ after $n$ appends. However, [resize] also does some writing, so in order to find a precise formula for $\nwritesfn(\varnitems)$ I need to analyze when [resize] is called. Consider the following lemma:

[lemma]
Let $\listname$ by a dynamic array. As $\varnitems$ are appended, the sequence of capacities $\listname$ takes on is $$S_C = 0,\ \initcapacity,\ \growthfactor\initcapacity,\ \growthfactor^2\initcapacity,\ \ldots\ \growthfactor^{\lceil \log_{\growthfactor} \varnitems - \log_{\growthfactor} \initcapacity \rceil}\initcapacity$$.

[proof]
The capacity starts out at zero and becomes $\initcapacity$ once one element is appended. From then on, it can only grow by a factor of $\growthfactor$ via [resize], so if $\growthfactor^i\initcapacity$ is the current capacity then $\growthfactor^{i + 1}\initcapacity$ must be the next capacity. By induction, the rest of the sequence is $$\left\{ \growthfactor^i\initcapacity \right\}_{i = 1}^k$$ for some $k$. Since the capacity is only as big as is needs to be, namely greater than or equal to $\varnitems$, $k$ is the smallest integer for which $\growthfactor^k\initcapacity \geq \varnitems$. Using this property, I now derive $k$.

@@

Each time [resize] is called, $i$ items are copied where $i$ is the current size of the dynamic array. Then the total number of items copied for $\varnitems$ appends is

@@

Finally, appending the writes made for every item in [append], I finally find an explicit formula for $\nwritesfn(\varnitems)$.

@@

I wish to find the space allocated while appending $\varnitems$ items to a dynamic array. I denote this quantity $\spacefn(\varnitems)$, and define it as the total length of all arrays allocated during $\varnitems$ [append] calls. I now derive a formula for $\spacefn(\varnitems)$.

From [lemma] I know the exact capacities $\listname$ takes on as $\varnitems$ items are appended. Dynamic arrays also have the property that $i \in S_C$ iff an array of length $i$ was allocated. Then the total length of arrays allocated is

### 2017-10-16 Draft2/Appending.Growth.tex
Rename some stuff

\funarrayimpl

@@

I again start off by finding the write cost when $\varnitems$ are added. I give the write cost function a slightly different name, $\nwritesnewfn(\varnitems)$, so I can compare it to $\nwritesfn(\varnitems)$ later. I also use $\growthfactornew$ and $\initcapacitynew$ instead of $\growthfactor$ and $\initcapacity$ for the growth factor and initial capacity, respectively.

[That lemma] still holds, since all of the assumptions made there are also true for funnel arrays. In particular, although funnel arrays resize slightly differently than dynamic arrays, the following claim is still true:

[other lemma]
The capacity of a funnel array grows by a constant factor $\growthfactornew$ after the first time it is called.

@@

For the base case, consider the second time [resize] is called: the current capacity is $\initcapacitynew$, and the next (total) capacity should be $\growthfactornew\initcapacitynew$. Then a buffer of size $(\growthfactornew - 1)\initcapacitynew$ should be added, and that is what the algorithm does.

@@

Now I take a look at [resize]. Unlike dynamic arrays, the number of writes performed does not equal $i$ if the current size is $i$. No items are being copied; the only source of writes is adding a buffer to the tail. The goal is to find the number of writes that causes.

@@

Let $k$ be the number of times the tail is added to. Since the tail is a dynamic array, the number of writes it makes is $\nwritesfn(k)$. Since [resize] is called $|S_R|$ times and it adds to the tail each time except for the first, $k = |S_R| - 1 = \lceil \log_{\growthfactornew} \varnitems - \log_{\growthfactornew} \initcapacitynew \rceil$. Then the formula for $\nwritesresizefn(n)$ is

@@

where $\initcapacity$ $\growthfactor$ are the capacity of the \textit{tail} (not the funnel array). I finally conclude that

@@

% Should this be C(n) instead of C?
The space function for $\varnitems$ [add] calls on a funnel array is denoted $\spacenewfn(\varnitems)$. Since funnel arrays never throw away item buffers, if the current capacity is $C$ then the total length of item buffers is also $C$. $\spacenewfn(\varnitems)$ is usually greater than $C$, however; this is because funnel arrays not only allocate item buffers, but stores references to them in the tail. Thus the space the tail allocates must also be determined.

I determined in [...] that if $\varnitems$ are added then the tail is added to $k$ times, where $k = \lceil \log_{\growthfactornew} \varnitems - \log_{\growthfactornew} \initcapacitynew \rceil$. Since the tail is a dynamic array, it allocates $\spacefn(k)$ space, where $\spacefn$ is the space function of the tail. Through a similar method to [...] it can be shown $\bigo(\spacefn(k)) = \bigo(k) = \bigo(\log \varnitems)$. Then % TODO: P-analysis should come after the explicit formula.

>>>>

\groarrayimpl

@@

I again start off by finding the write cost when $\varnitems$ are appended. I give the write cost function a slightly different name, $\nwritesnewfn(\varnitems)$, so I can compare it to $\nwritesfn(\varnitems)$ later. I also use $\growthfactornew$ and $\initcapacitynew$ instead of $\growthfactor$ and $\initcapacity$ for the growth factor and initial capacity, respectively.

[That lemma] still holds, since all of the assumptions made there are also true for growth arrays. In particular, although growth arrays resize slightly differently than dynamic arrays, the following claim is still true:

[other lemma]
The capacity of a growth array grows by a constant factor $\growthfactornew$ after the first time it is called.

@@

For the base case, consider the second time [resize] is called: the current capacity is $\initcapacitynew$, and the next (total) capacity should be $\growthfactornew\initcapacitynew$. Then a buffer of size $(\growthfactornew - 1)\initcapacitynew$ should be appended, and that is what the algorithm does.

@@

Now I take a look at [resize]. Unlike dynamic arrays, the number of writes performed does not equal $i$ if the current size is $i$. No items are being copied; the only source of writes is appending a buffer to the tail. The goal is to find the number of writes that causes.

@@

Let $k$ be the number of times the tail is appended to. Since the tail is a dynamic array, the number of writes it makes is $\nwritesfn(k)$. Since [resize] is called $|S_R|$ times and it appends to the tail each time except for the first, $k = |S_R| - 1 = \lceil \log_{\growthfactornew} \varnitems - \log_{\growthfactornew} \initcapacitynew \rceil$. Then the formula for $\nwritesresizefn(n)$ is

@@

where $\initcapacity$ $\growthfactor$ are the capacity of the \textit{tail} (not the growth array). I finally conclude that

@@

% Should this be C(n) instead of C?
The space function for $\varnitems$ [append] calls on a growth array is denoted $\spacenewfn(\varnitems)$. Since growth arrays never throw away item buffers, if the current capacity is $C$ then the total length of item buffers is also $C$. $\spacenewfn(\varnitems)$ is usually greater than $C$, however; this is because growth arrays not only allocate item buffers, but stores references to them in the tail. Thus the space the tail allocates must also be determined.

I determined in [...] that if $\varnitems$ are appended then the tail is appended to $k$ times, where $k = \lceil \log_{\growthfactornew} \varnitems - \log_{\growthfactornew} \initcapacitynew \rceil$. Since the tail is a dynamic array, it allocates $\spacefn(k)$ space, where $\spacefn$ is the space function of the tail. Through a similar method to [...] it can be shown $\bigo(\spacefn(k)) = \bigo(k) = \bigo(\log \varnitems)$. Then % TODO: P-analysis should come after the explicit formula.

### 2017-10-16 Draft2/CopyArray.Blurb.tex
Rename some stuff

Users often want to take list structures, such as dynamic arrays, and convert them into plain arrays. There are multiple reasons why someone would want to do this after they are done adding to the list:

\begin{itemize}
	\item Plain arrays hold on to exactly the amount of memory needed to hold their elements. However, dynamic and funnel arrays allocate more space than necessary to optimize adding new items.
	\item The user wants to call a function in third-party code that takes a plain array as an argument.
	\item
	\item Plain arrays are contiguous, while funnel arrays are fragmented and have worse locality.
	\item The indexer of funnel arrays is several times slower than that of plain arrays, whether the $\bigo(1)$ or $\bigo(\log \varnitems)$ implementation is chosen.
\end{itemize}

>>>>

Users often want to take list structures, such as dynamic arrays, and convert them into plain arrays. There are multiple reasons why someone would want to do this after they are done appending to the list:

\begin{itemize}
	\item Plain arrays hold on to exactly the amount of memory needed to hold their elements. However, dynamic and growth arrays allocate more space than necessary to optimize appending new items.
	\item The user wants to call a function in third-party code that takes a plain array as an argument.
	\item
	\item Plain arrays are contiguous, while growth arrays are fragmented and have worse locality.
	\item The indexer of growth arrays is several times slower than that of plain arrays, whether the $\bigo(1)$ or $\bigo(\log \varnitems)$ implementation is chosen.
\end{itemize}
